---
title: "Sunrise and sunset in Vietnam"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r include = FALSE, eval = FALSE}
# The code that adds the .nojekyll file to the GitHub repository:
file <- ".nojekyll"
file.create(file)
gert::git_add(file)
gert::git_commit("Adding the .nojekyll file")
gert::git_push()
rm(file)
```

```{r include = FALSE, eval = FALSE}
run_all <- function() {
  knitr::purl("index.qmd", "tmp.R", documentation = FALSE)
  source("tmp.R")
  file.remove("tmp.R")  
}

rerun_all <- function() {
  rm(list  = grep("run_all", ls(envir = .GlobalEnv), value = TRUE, invert = TRUE),
     envir = .GlobalEnv)
  run_all()
}
```

```{r include = FALSE, purl = FALSE}
setHook("plot.new", NULL, action = "replace")

par2 <- function(...) par(..., mgp = c(1.5, .5, 0), bty = "n")

knitr::knit_hooks$set(
  margin = function(before, options, envir) {
    if (before) par2(plt = c(.105, .97, .15, .95)) else NULL
  })

figw <- 7  # default is 7 and maximum possible is 8.3
figh <- 5  # default is 5
eps <- 1

knitr::opts_chunk$set(margin     = TRUE,
                      fig.retina = 2,
                      fig.align  = "center",
                      fig.width  = eps * figw,
                      fig.height = eps * figh) 
```


Whereas the durations of day and night as a function of time follow a sinusoid
function, this is not the case for the timing of the sunrise and sunset because of the
facts (i) that the orbit of the earth around the sun is an ellipse and not a circle and
(ii) the rotation axis of the earth is tilted on the orbital plan. Here we illustrate
all this in a given location.

```{r message = FALSE}
library(hms)
library(dplyr)
library(suncalc)
library(lubridate)
library(purrr)
library(sf)
```

Computing the sunrise and sunset times for Saigon for 2023 and 2024:

```{r}
timings <- seq.Date(ymd(20230101), ymd(20241231), 1) |> 
  getSunlightTimes(lat = 10.802506629026347, lon = 106.73305962858632) |> # Saigon
  select(date, sunrise, sunset) |>  
  mutate(across(-date, ~ .x + 7 * 60 * 60)) |>  # UTC to ICT transformation
  mutate(across(-date, as_hms)) |> 
  mutate(day_duration = sunset - sunrise)
```

Looking at timings of earliest and latest sunrises and sunsets:

```{r}
timings |> 
  arrange(sunrise) |> 
  head()

timings |> 
  arrange(desc(sunrise)) |> 
  head()

timings |> 
  arrange(sunset) |> 
  head()

timings |> 
  arrange(desc(sunset)) |> 
  head()
```

A tuning of the `plot()` function:

```{r}
plot2 <- function(...) plot(..., type = "l", lwd = 2)
```

The duration of the day as a function of time:

```{r}
with(timings, plot2(date, day_duration / 60 / 60, col = 2, xlab = "",
                    ylab = "day duration (hours)"))
abline(v = ymd(20240101))
```

Timing of sunrise and sunset as a function time:

```{r}
with(timings, {
  plot2(date, sunrise / 60 / 60, col = 2, ylim = c(5, 19),
        ylab = "time of daylight")
  lines(date, sunset / 60 / 60, col = 4, lwd = 2)
})
abline(v = ymd(20240101))
```

An alternative to the previous plot:

```{r}
opar <- par(mfrow = c(2, 1), plt = c(.1, .99, .12, 1))

with(timings, plot2(date, sunrise / 60 / 60, col = 2, xlab = "",
                    ylab = "time of sunrise", yaxt = "n"))
ats <- paste(rep(c("05", "06"), each = 6),
             rep(sprintf("%02d", 10 * 0:5), 2), sep = ":")[4:8]
axis(2, at = time_length(hm(ats), "hours"), ats)
abline(v = ymd(20240101))

with(timings, plot2(date, sunset / 60 / 60, col = 4, xlab = "",
                    ylab = "time of sunset", yaxt = "n"))
ats <- paste(rep(c("17", "18"), each = 6),
             rep(sprintf("%02d", 10 * 0:5), 2), sep = ":")[4:9]
axis(2, at = time_length(hm(ats), "hours"), ats)
abline(v = ymd(20240101))

par(opar)
```

A function that calculates the timings of sunrise and sunset as a function of time (in
2023 and 2024) for a given location in the ICT time zone:

```{r}
calculate_timings <- function(lat = 10.802506629026347, lon = 106.73305962858632) {
  seq.Date(ymd(20230101), ymd(20241231), 1) |> 
    getSunlightTimes(lat = lat, lon = lon) |>
    select(date, sunrise, sunset) |>  
    mutate(across(-date, ~ .x + 7 * 60 * 60)) |> 
    mutate(across(-date, as_hms)) |> 
    mutate(day_duration = sunset - sunrise)
}
```

Computing the longitudes and latitudes of the barycenters of the provinces of Vietnam:

```{r}
gadm1 <- readRDS("~/Desktop/gadm36_VNM_1_sf.rds")
st_crs(gadm1) <- st_crs(gadm1)

timings_list <- gadm1 |>
  mutate(centroid  = geometry |> st_centroid() |> st_coordinates(),
         latitude  = centroid[, "Y"],
         longitude = centroid[, "X"]) |> 
  st_drop_geometry() |> 
  select(latitude, longitude) |> 
  with(map2(latitude, longitude, calculate_timings))
```

Plotting the timing of sunrise and sunset as a function of time for the 63 provinces of
Vietnam:

```{r}
opar <- par(mfrow = c(2, 1), plt = c(.1, .99, .12, 1))

with(timings, plot(date, sunrise / 60 / 60, type = "n", xlab = "",
                   ylab = "time of sunrise", yaxt = "n", ylim = c(5, 7)))
ats <- paste(rep(c("05", "06"), each = 3),
             rep(sprintf("%02d", seq(0, 40, 20)), 2), sep = ":")
axis(2, at = time_length(hm(ats), "hours"), ats)
abline(v = ymd(20240101))
walk(timings_list, ~ with(.x, lines(date, sunrise / 60 / 60, col = 2)))

with(timings, plot(date, sunset / 60 / 60, type = "n", xlab = "",
                   ylab = "time of sunset", yaxt = "n", ylim = c(17, 19)))
ats <- paste(rep(c("17", "18"), each = 3),
             rep(sprintf("%02d", seq(0, 40, 20)), 2), sep = ":")
axis(2, at = time_length(hm(ats), "hours"), ats)
abline(v = ymd(20240101))
walk(timings_list, ~ with(.x, lines(date, sunset / 60 / 60, col = 4)))

par(opar)
```

